Clause size reduction with $i$-UIP Learning

### 2020-02-12 paper.tex
more fb changes

\documentclass[runningheads]{llncs}
\usepackage{times}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
%\usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
\usepackage{xspace}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{array}
\usepackage{xcolor}
\usepackage{theorem}
\usepackage[T1,hyphens]{url}

\begin{document}
\title{$i$-UIP Clause Learning}
\author{Nick Feng \and Fahiem Bacchus}
\authorrunning{N. Feng and F. Bacchus}
\institute{Department of Computer Science, University of Toronto, Canada, 
  \email{\{nfeng,fbacchus\}@cs.toronto.edu}}

\newcommand{\sat}{SAT\xspace}
\newcommand{\code}[1]{\text{#1}}
\newcommand{\inputF}{\ensuremath{F}}
\newcommand{\oneUIPClause}{\ensuremath{C_{1}}}
\newcommand{\iUIPClause}{\ensuremath{C_{i}}}
\newcommand{\assertionTrail}{\ensuremath{\phi}}
\newcommand{\LBD}{\text{LBD}}
\newcommand{\GAP}{\text{Gap}}
\newcommand{\IUIP}{\textbf{$i$-UIP}}
\newcommand{\IUIPPURE}{\textbf{Pure-i-UIP}}
\newcommand{\IUIPMIN}{\textbf{Min-i-UIP}}
\newcommand{\IUIPGreedy}{\textbf{i-UIP-Greedy}}
\newcommand{\IUIPActive}{\textbf{i-UIP-Inclusive}}
\newcommand{\IUIPDist}{\textbf{i-UIP-Exclusive}}
\newcommand{\MapleBase}{\textit{MapleCOMSPS\_LRB}}
\newcommand{\MapleSeven}{\textit{MapleLCMDist}}
\newcommand{\MapleNine}{\textit{MapleLCMDiscChronoBT-DL-v3} }
\newcommand{\MapleEight}{\textit{MapleLCMDiscChronoBT}}
\newcommand{\MapleNineShort}{\textit{MapleCB-DL} }
\newcommand{\expSAT}{\textit{expMaple\_CM\_GCBumpOnlyLRB} }
\newcommand{\expSATShort}{\textit{expMaple} }
\newcommand{\MapleIUIPPURE}{\text{Maple-\IUIPPURE}}
\newcommand{\MapleIUIMIN}{\text{Maple-\IUIPMIN}}
\newcommand{\MapleEightShort}{\textit{MapleCB}}
\newcommand\Set[2]{\{\,#1\mid#2\,\}}
\newcommand\Array[2]{[\,#1\mid#2\,]}
\newcommand\target{\textbf{Succ_t}}
\newcommand\trail{\ensuremath{\mathcal{T}}}
\newcommand\trailIdx[1]{\ensuremath{\iota(#1)}}
\newcommand\var{\text{var}}
\newcommand\true{\textsc{true}\xspace}
\newcommand\false{\textsc{false}\xspace}
\newcommand\reason{\textit{reason}}
\setlength{\theorempreskipamount}{3pt}
\setlength{\theorempostskipamount}{3pt}
\newtheorem{Obs}{Observation}
\newtheorem{Cor}{Corollary}
\newtheorem{defn}{Definition}
\newtheorem{thm}{Theorem}

@@

\maketitle              % typeset the header of the contribution
% 
\begin{abstract}
\end{abstract}

\section{Introduction}
Clause learning is an essential technique in \sat solvers. And there
is good empirical evidence that clause learning is, in fact, the most
important technique used in modern \sat solvers
\cite{DBLP:conf/sat/KatebiSS11}. In early \sat research a number of
different clause learning techniques were proposed
\cite{DBLP:conf/iccad/ZhangMMM01,DBLP:conf/iccad/SilvaS96,DBLP:journals/tc/Marques-SilvaS99,DBLP:conf/aaai/BayardoS97}.
However, following the revolutionary performance improvements achieved
by the Chaff \sat solver, the field has converged on using the 1-UIP
(first Unique Implication Point) scheme
\cite{DBLP:conf/iccad/ZhangMMM01} employed in Chaff
\cite{DBLP:conf/dac/MoskewiczMZZM01} (as well as other techniques
pioneered in the Chaff solver).\footnote{The idea of UIP clauses was
  first mentioned in \cite{DBLP:journals/tc/Marques-SilvaS99}, and
  1-UIP clauses along with other UIP clauses were learnt and used in
  the earlier GRASP \sat solver.} Since then almost all \sat solvers
have employed the 1-UIP clause learning scheme, along with clause
minimization \cite{DBLP:conf/sat/SorenssonB09}, as their primary
method for learning new clauses.

However, as this early work showed, other clause learning schemes can
be used in SAT solvers without changes to their main data
structures. Furthermore, advances in our understanding allow us to
better understand the potential advantages and disadvantages of these
alternate schemes. In this paper we reexamine some of these previous
clause learning schemes, with a focus on the $i$-UIP schme described
in \cite{DBLP:conf/iccad/ZhangMMM01}. With current ideas we can see
that in its original form $i$-UIP clause learning suffers a
significant disadvantages over 1-UIP clause learning. We present a
number of different methods to impove $i$-UIP clause learning so as to
over come these disadvantages. We show that these different methods
for $i$-UIP clause learning can provide useful performance gains over
the standard 1-UIP scheme.

\section{Background}
We assume the reader has some familiarity with Conjunctive Normal Form
(CNF) formulas, conflict-driven clause learning (CDCL) \sat solvers
\cite{DBLP:series/faia/SilvaLM09}, unit-propagation, and conflict
analysis.

\begin{enumerate}
\item Trail; Reasons and their properties; decision levels, trail order.
\item conflict clause. Trail has the property that conflict exists
    only at the deepest level.
\item Clauses derived by trail resolution. Identified by variables
    appearing on the trail.
\item No point in doing this in non-trail order.
\item a specific case of trivial resolutions.
\item Clause derived by trail resolution iff it is a cut in the
    implication graph.
\item Note that all implemented clause learning techniques use
    sequences of trail resolutions. In particular, they all start with
    a specific conflict and resolve only with reason clauses.
    This includes clause minimization.
\item Asserting clause---one lit at deepest level of trail. Only
    asserting clauses are gauranteed to be new.
\item 1-UIP is asserting clause generated with fewest trail resolution
    steps. Each trail resolution step can only increase the number of
    levels in the clause. So LBD is minimized by 1-UIP clauses.
\item Description of i-uip, all-uip, all decisions (useful in some cases)
\end{enumerate}

CDCL \sat solvers maintain a \textbf{trail} which is a sequence of
literals that have been assigned \true by the solver. Literals newly
assigned \true are placed at the end of the trail, and literals are
always unassigned and removed from the end of the trail first;
i.e. the trail acts as a stack.

Let $\trail$ be the trail. For any literal $\ell$ we let
$\trailIdx{\ell}$ be the index of $\ell$ in $\trail$. That is,
$\trail[\trailIdx{\ell}] = \ell$. If $\ell$ is not in the trail, then
$\trailIdx{\ell} = \varnothing$, and we use the convention that
$\trail[\varnothing] = \varnothing$. The literals on \trail are
sequenced in the order they were assigned \true. Hence
$\trailIdx{\ell} < \trailIdx{\ell'}$ implies that $\ell$ was assigned
\true before $\ell'$. When $\trailIdx{\ell} < \trailIdx{\ell'}$ we say
that $\ell$ \textit{appears above} $\ell'$ on the trail. We use the
convention that $\trailIdx{\ell} < \trailIdx{\ell'}$ whenever $\ell$
is in the trail and $\ell'$ is not; i.e., $i < \varnothing$ for any
index $i$.

The literals on $\trail$ have been assigned \true by the solver by
either (a) having been set as a decisions or (b) having forced by
unit-propagation. If $\ell\in \trail$ was was set by unit-propgation
it will have an associated reason clause $\reason(\ell)$. This clause
has two properties (1) $\ell\in \reason(\ell)$ and (2) for every
$x\in \reason(\ell)$ we have that
$\trailIdx{\lnot x} < \trailIdx{\ell}$. That is, $\reason(\ell)$ is a
clause that implies that $\ell$ must assigned \true because every
other literal in it has been assigned \false above $\ell$ in
$\trail$. Literals $\ell$ set as decsions do not have reason clauses
and we let $\reason(\ell) = \varnothing$.

CDCL \sat solvers perform clause learning when the literals on the
trail falsify a clause $c$. '

\section{Find i-uip Clause}
\label{sec:i-uip}
Classical CDCL SAT solvers analyze each conflict with 1-UIP resolution
learning scheme to derive an asserting clause $\oneUIPClause$. Even
though prior studies shows that further resolutions on $\oneUIPClause$
against the assertion trail $\assertionTrail$ cannot reduce the
clause's literal block distance ($\LBD$), the resolutions could
potentially reduce the clause's size. Clause size is an important
quality measurement because a smaller clause 1) consumes less memory,
2) requires less steps to force a literal and 3) decreases the size of
future conflict clauses.

Inspired by both $\LBD$ and clause size, i-UIP resolution learning
attempts to resolve away literals in $\oneUIPClause$ against the
assertion trail $\assertionTrail$ to minimize the clause size without
increasing the clause's $\LBD$. The goal of i-UIP learning is to find
a clause $\iUIPClause$ whose literals are either the unique
implication points in their respective decision levels or are directly
implied by literals from a foreign decision
level. Alg. ~\ref{alg:i-uip} is a pseudo-code implementation of i-UIP
learning.

\begin{algorithm}[t]
\caption{\IUIP}\label{alg:i-uip}
\begin{algorithmic}[1]
\Require  $\oneUIPClause$ is a valid and minimized 1-UIP clause
\Require  $\assertionTrail$ is a valid assertion trail
\Procedure{\IUIP}{$\oneUIPClause, \assertionTrail$} 
    \State $\iUIPClause \gets \oneUIPClause $ \Comment{initialize i-UIP clause} \label{ln:init}
    \State \code{DecisionLvs} $\gets$ Decision levels in $\oneUIPClause$ in descending order
\For{$i \in \code{DecisionLvs}$} \label{ln:iterDecisionLvls}
    \State $L_{i} \gets \Set{l}{\code{level}(l) = i \wedge l \in \iUIPClause}$
    \While{$|\code{unmarked}(L_i)| > 1$} \label{ln:uip}
        \State $p \gets \text{lit with the highest trail position in } L_{i} $ \label{ln:trailaccess}
        \If{$\exists q \in \code{Reason}(\neg{p}, \assertionTrail) \cdot \code{level}(q) \not\in$ \code{DecisionLvs}} \label{ln:resolvable}
            \If{$\IUIPPURE$}
                \State Unresolve literals at level $i$  \label{ln:unresolve}
                \State Go to the next decision level $i$ \label{ln:abondon}
            \ElsIf{$\IUIPMIN$}
                \State $\code{Mark}(p)$ \label{ln:mark}
            \EndIf
        \Else 
            \State  $\iUIPClause \gets (\oneUIPClause ) \bowtie \code{Reason}(\neg{p}, \assertionTrail)$
           \State  $\code{Update}(L_{i})$
        \EndIf
    \EndWhile
\EndFor \label{ln:endCompleteResolution}
\\
\If{$\IUIPPURE$}\label{ln: 2ndMinStart}   \Comment{$\iUIPClause$ from $\IUIPPURE$ can be minimized}
    \State $\iUIPClause  \gets \textbf{Minimize}(\iUIPClause)$ 
\EndIf\label{ln: 2ndMinEnd}
\\
\If{$|\iUIPClause| < |\oneUIPClause|$} \label{ln:checkSize}
    \State \textbf{return} $\iUIPClause$
\Else
    \State \textbf{return} $\oneUIPClause$
\EndIf  \label{ln:cSizeCheck}
\EndProcedure
\end{algorithmic}
\end{algorithm}

The algorithm $\IUIP$ computes $\iUIPClause$ from (line~\ref{ln:init}
to \ref{ln:endCompleteResolution}), and then returns the smaller
clause between $\oneUIPClause$ and $\iUIPClause$
(line~\ref{ln:checkSize} to \ref{ln:cSizeCheck}).  The algorithm first
initializes $\iUIPClause$ with the minimized\cite{}
$\oneUIPClause$. Next, the algorithm iterate through the decision
levels of $\iUIPClause$ in descending order
(line~\ref{ln:iterDecisionLvls}) and tries to find the unique
implication point in each level (line~\ref{ln:uip}). Since $\IUIP$
needs to preserve $\iUIPClause$'s $\LBD$ during resolutions, before
resolving away a target literal $p$, the algorithm preemptively checks
whether the reasoning clause of $\neg{p}$ contains any literal $q$
from an unseen decision level (line~\ref{ln:resolvable}). If such a
literal $q$ exists, then the algorithm cannot find the unique
implication point (UIP) at the current level $i$ because resolving
away $p$ will introduce $q$ into the clause and increases $\LBD$. One
solution is to abandon level $i$ by reverting back to the state before
any literal at level $i$ is resolved away (line~\ref{ln:unresolve}),
and then move on to the next level (line~\ref{ln:abondon}). We denote
$\IUIP$ learning with this solution as $\IUIPPURE$. The final
$\iUIPClause$ produced by $\IUIPPURE$ will contains exactly one
literal for levels where the LBD-persevering UIP exist. However,
$\IUIPPURE$ does not minimize the number of literals in the decision
levels where the LBD-preserving UIP does not exist, and clause
minimization techniques\cite{} can be used to further reduce the
clause size (line~\ref{ln: 2ndMinStart} to \ref{ln: 2ndMinEnd}).

When $\IUIP$ cannot resolve away a literal $p$ without increasing
$\iUIPClause$'s LBD, instead of skipping the entire decision level
$i$, another solution $\IUIPMIN$ simply marks literal $p$ and keep it
in $\iUIPClause$ (line~\ref{ln:mark}). The learning scheme then
continue the resolutions at level $i$. $\IUIPMIN$ will ignore all the
unsolvable literals and find the "local" unique implication point at
every decision level.  The algorithm terminates when there is exactly
one unmarked literal left for each decision level of $\iUIPClause$,
representing the local unique implication points. The clause
$\iUIPClause$ learnt by $\IUIPMIN$ is minimized\footnote{clause size
  cannot be reduced via minimization techniques\cite{}.}, we defer the
proof of minimization to Appendix~\ref{}.

\begin{example}
    Consider the assertion trail
    $\assertionTrail =
    {..a_2,b_2,c_2,d_2..e_5,f_5,g_5,h_6,i_6,j_6,k_6..}$ where literals
    $a_2 ... k_{6}$ are asserted in order with decision level shown as
    the underscript. Let $L_{ext}$ denote the set of literals on
    $\assertionTrail$ outside of decision level 2,5,6 and 10, and
    $l^{*}_{ext}$ be some literal in $L_{ext}$. The reasoning clauses
    involved in building $\assertionTrail$ are:
\begin{enumerate}
  \item[] $C_k = \{\neg{f_5} \vee \neg{j_6} \vee k_6\}$,
  $C_j = \{\neg{f_5} \vee \neg{i_6} \vee j_6\}$, 
  $C_i = \{\neg{e_5} \vee \neg{h_6} \vee i_6\}$,
  $C_g = \{\neg{a_2} \vee \neg{f_5} \vee g_5\}$
  $C_f = \{\neg{l^{*}_{ext}} \vee \neg{e_5} \vee f_5\}$,
  $C_d = \{\neg{b_2} \vee \neg{c_2} \vee d_2\}$,
  $C_c = \{\neg{a_2} \vee \neg{b_2} \vee c_2 \}$,
  $C_b = \{\neg{l^{*}_{ext}} \vee \neg{a_2} \vee b_2\}$
\end{enumerate}
Suppose 1-UIP learning produces clause
$\oneUIPClause = \{\neg{m_{10}} \vee \neg{k_6} \vee \neg{j_6} \vee
\neg{i_6} \vee \neg{h_6} \vee \neg{g_5} \vee \neg{d_2} \vee
\neg{c_2}\}$ where $\neg{k_{10}}$ is the UIP from the conflicting
level. $\IUIP$ first tries to find the UIP for level 6 by resolving
$\oneUIPClause$ with $C_k$, $C_j$ and, $C_i$ in order and produces
clause
$\iUIPClause = \{\neg{m_{10}} \vee \neg{h_6} \vee \neg{g_5} \vee
\neg{f_5} \vee \neg{e_5} \vee \neg{d_2} \vee \neg{c_2} \}$ where
$\neg{h_6}$ is the UIP.

$\IUIP$ attempts to find the UIP for level 5 by resolving $C_i$ with
$C_g$ and $C_f$ in order. However, resolving with $C_f$ will introduce
$\neg{l^{*}_{ext}}$ into $C_i$ and increase the clause's
LBD. $\IUIPPURE$ undoes all resolutions at level 5 and skips the
entire level, whereas $\IUIPMIN$ only skips the resolution with $C_f$
and tries to find the UIP by ignoring $\neg{f_5}$.  After level 5,
$\IUIPPURE$ preserves the same clause and $\IUIPMIN$ produces
$\iUIPClause = \{\neg{m_{10}} \vee \neg{h_6} \vee \neg{f_5} \vee
\neg{e_5} \vee \neg{d_2} \vee \neg{c_2} \vee \neg{a_2}\}$ where
$\neg{e_5}$ is the UIP when $\neg{f_5}$ is ignored. The same $\IUIP$
procedure is then performed on level 2 to produce the final
$\iUIPClause$.

>>>>

\documentclass[runningheads]{llncs}
\usepackage{times}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
%\usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
\usepackage{xspace}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{array}
\usepackage{xcolor}
\usepackage{theorem}
\usepackage[T1]{fontenc}
% \usepackage[T1,hyphens]{url}
\usepackage{hyperref}
%\PassOptionsToPackage{hypnens}{url}
\title{Clause size reduction with $i$-UIP Learning}
\author{Nick Feng \and Fahiem Bacchus}
\authorrunning{N. Feng and F. Bacchus}
\institute{Department of Computer Science, University of Toronto, Canada, 
  \email{\{nfeng,fbacchus\}@cs.toronto.edu}}

\algrenewcommand\algorithmicprocedure{}
\algrenewcommand\algorithmicthen{}
\algrenewcommand\algorithmicdo{}

\newcommand{\fs}[1]{\fontsize{#1}{#1}\selectfont}
\newcommand{\fss}[2]{\fontsize{#1}{#2}\selectfont}
\newcommand{\sat}{SAT\xspace}
\newcommand{\code}[1]{\text{#1}}
\newcommand{\inputF}{\ensuremath{F}}
\newcommand{\oneUIPClause}{\ensuremath{C_{1}}}
\newcommand{\iUIPClause}{\ensuremath{C_{i}}}
\newcommand{\assertionTrail}{\trail}
\newcommand{\LBD}{\text{LBD}}
\newcommand{\GAP}{\text{Gap}}
\newcommand{\IUIP}{\textbf{LBDstable-$i$UIP}}
\newcommand{\tryuiplevel}{\textbf{try-uip-level}\xspace}
\newcommand{\IUIPPURE}{\textbf{Pure-$i$UIP}}
\newcommand{\IUIPMIN}{\textbf{Min-$i$UIP}}
\newcommand{\IUIPGreedy}{\textbf{i-UIP-Greedy}}
\newcommand{\IUIPActive}{\textbf{i-UIP-Inclusive}}
\newcommand{\IUIPDist}{\textbf{i-UIP-Exclusive}}
\newcommand{\MapleBase}{\textit{MapleCOMSPS\_LRB}}
\newcommand{\MapleSeven}{\textit{MapleLCMDist}}
\newcommand{\MapleNine}{\textit{MapleLCMDiscChronoBT-DL-v3} }
\newcommand{\MapleEight}{\textit{MapleLCMDiscChronoBT}}
\newcommand{\MapleNineShort}{\textit{MapleCB-DL} }
\newcommand{\expSAT}{\textit{expMaple\_CM\_GCBumpOnlyLRB} }
\newcommand{\expSATShort}{\textit{expMaple} }
\newcommand{\MapleIUIPPURE}{\text{Maple-\IUIPPURE}}
\newcommand{\MapleIUIMIN}{\text{Maple-\IUIPMIN}}
\newcommand{\MapleEightShort}{\textit{MapleCB}}
\newcommand{\Set}[2]{\{\,#1\mid#2\,\}}
\newcommand{\Array}[2]{[\,#1\mid#2\,]}
\newcommand{\target}{\textbf{Succ_t}}
\newcommand{\trail}{\ensuremath{\mathcal{T}}}
\newcommand{\trailIdx}[1]{\ensuremath{\iota(#1)}}
\newcommand{\dlevel}[1]{\ensuremath{\mathit{decLvl}(#1)}}
\newcommand{\dlevels}{\ensuremath{\mathit{decLvls}}}
\newcommand{\var}{\text{var}}
\newcommand{\true}{\textsc{true}\xspace}
\newcommand{\false}{\textsc{false}\xspace}
\newcommand{\reason}[1]{\ensuremath{\mathit{reason}(#1)}}
\newcommand{\resolve}{\bowtie}
\renewcommand{\implies}{\rightarrow}
\newcommand{\ctry}{C_{\mathit{try}}}
\newcommand{\type}{\textit{type}}
\newcommand{\unmarked}{\textbf{unmarked}}
\newcommand{\mrk}{\textbf{mark}}
\setlength{\theorempreskipamount}{3pt}
\setlength{\theorempostskipamount}{3pt}

\newtheorem{Obs}{Observation}
\newtheorem{Cor}{Corollary}
\newtheorem{defn}{Definition}
\newtheorem{thm}{Theorem}

@@

\begin{document}
\maketitle              % typeset the header of the contribution
% 
\begin{abstract}
\end{abstract}

\section{Introduction}
Clause learning is an essential technique in \sat solvers. There is good
evidence to indicate that it is, in fact, the most important technique
used in modern \sat solvers \cite{DBLP:conf/sat/KatebiSS11}. In early
\sat research a number of different clause learning techniques were
proposed
\cite{DBLP:conf/iccad/ZhangMMM01,DBLP:conf/iccad/SilvaS96,DBLP:journals/tc/Marques-SilvaS99,DBLP:conf/aaai/BayardoS97}.
However, following the revolutionary performance improvements achieved
by the Chaff \sat solver, the field has converged on using the 1-UIP
(first Unique Implication Point) scheme
\cite{DBLP:conf/iccad/ZhangMMM01} employed in Chaff
\cite{DBLP:conf/dac/MoskewiczMZZM01} (as well as other techniques
pioneered in the Chaff solver).\footnote{The idea of UIP clauses was
  first mentioned in \cite{DBLP:journals/tc/Marques-SilvaS99}, and
  1-UIP clauses along with other UIP clauses were learnt and used in
  the earlier GRASP \sat solver.} Since then almost all \sat solvers
have employed the 1-UIP clause learning scheme, along with clause
minimization \cite{DBLP:conf/sat/SorenssonB09}, as their primary
method for learning new clauses.

However, other clause learning schemes can be used in SAT solvers
without changes to their main data structures. Furthermore, advances
in our understanding allow us to better understand the potential
advantages and disadvantages of these alternate schemes. In this paper
we reexamine some of these previous clause learning schemes, with a
focus on the schemes described in
\cite{DBLP:conf/iccad/ZhangMMM01}. Improved understanding of \sat
solvers, obtained from the last decade of research, allows us to see
that in their original form these other clause learning schemes suffer
significant disadvantages over 1-UIP clause learning.

One of the previously proposed schemes was the $i$-UIP scheme
\cite{DBLP:conf/iccad/ZhangMMM01}. In this paper we propose a new way
to exploit the main ideas of this scheme that avoids its main
disadvantages. In particular, we propose to use a $i$-UIP like clause
learning scheme to generate smaller learnt clauses which retain the
good properties of standard 1-UIP clauses. Our method is related to,
but not the same as various clause mininimization methods that aim to
remove literals from the 1-UIP clause, e.g.,
\cite{DBLP:conf/sat/SorenssonB09,DBLP:conf/ijcai/LuoLXML17,DBLP:conf/sat/WieringaH13}.
Our method is orthogonal to clause minimization. In particular, 
our approach can learn a clause that is completely different from the
1-UIP clause but which still serves all of the same purposes as the 1-UIP
clause. Like the 1-UIP clause, various minimization techniques can be
applied on top of our method to further reduce the size of the clauses
we learn.

We present various versions of our method and show that these variants
are often capable of learning shorter clauses that the 1-UIP scheme,
and that this can lead to useful performance gains in state of the art
\sat solvers.

\section{Clause Learning Framework}
We first provide some background and a framework for understanding
clause learning as typically used in CDCL \sat solvers. A
propositional formula $F$ expressed in Conjunctive Normal Form (CNF),
contains a set of variables $V$. A literal is a variable $v\in V$ or
its negation $\lnot v$. For a literal $\ell$ we let $\var(\ell)$
denote its underlying variable. A CNF consists of a conjunction of
clauses, each if which is a disjunction of literals. We often view a
clause as being a set of literals and employ set notation, e.g.,
$\ell\in C$ and $C'\subset C$.

Two clauses $c_1$ and $c_2$ can be \textit{resolved} when they contain
conflicting literals $\ell\in c_1$ and $\lnot \ell \in c_2$. Their
resolvant $c_1 \resolve c_2$ is the new clause
$(c_1 \cup c_2) - \{\ell, \lnot \ell\}$. The resolvant is tautological
if $c_1$ and $c_2$ contain more than one pair of conflicting literals.

We assume the reader is familiar with the operations of CDCL \sat
solvers, and the main data structures used in such solvers. A good
source for this background is \cite{DBLP:series/faia/SilvaLM09}.

\paragraph{The Trail.}
CDCL \sat solvers maintain a \textbf{trail}, $\trail$, which is a
\textit{non-contradictory, non-redundant sequence of literals} that
have been assigned \true by the solver; i.e.
$\ell\in\trail \implies \lnot\ell \not\in\trail$, and $\trail$
contains no duplicates. Newly assigned literals are added to the end
of the trail, and on backtrack literals are removed from the end of
the trail and unassigned.  If literal $\ell$ is on the trail let
$\trailIdx{\ell}$ denote its index on the trail sequence
($\trail[\trailIdx{\ell}] = \ell$). For convenience, we also let
$\trailIdx{\ell} = \trailIdx{\lnot \ell} = \trailIdx{\var(\ell)}$,
where $\var(\ell)$ is $\ell$'s variable, even though $\lnot \ell$ nor
$\var(\ell)$ are actually on $\trail$. If $x$ and $y$ are both on the
trail and $\trailIdx{x} < \trailIdx{y}$ we say that \textit{$x$
  appears before $y$ on the trail}. In a slight abuse of notation we
also use $\trailIdx{v}$ on a variable $v$ where
$\trailIdx{v}=\trailIdx{\ell}$ and $\ell$ a literal of $v$ on the
trail.

Two types of true literals appear on the trail: decision literals that
have been assumed to be true by the solver, and unit propagated
literals that are forced to be true because they are the sole remining
unfalsified literal of a clause. Each literal $\ell\in\trail$ has a
decision level $\dlevel{\ell}$ which is equal to the number of
decision literals appearing before $\ell$ on the trail plus one;
hence, $\dlevel{d}=1$ for the first decision literal $d\in\trail$.
The set of literals on $\trail$ that have the same decision level
forms a contiguous subsequence of $\trail$ that starts with a decision
literal $d_i$ and ends just before the next decision literal
$d_{i+1}$. If $\dlevel{d_i} = i$ we call this subsequence of $\trail$
the \textit{$i$-th decision level}.

Each literal $\ell\in\trail$ also has a clausal reason
$\reason{\ell}$. If $\ell$ is a unit propgated literal,
$\reason{\ell}$ is a clause of the formula such that
$\ell \in \reason{\ell}$ and
$\forall x \in \reason{\ell}.\, x\neq \ell \implies (\lnot x \in
\trail \land \trailIdx{\lnot x} < \trailIdx{\ell}$. That is,
$\reason{\ell}$ is a clause that has become unit implying $\ell$ due
to the literals on the trail above $\ell$. If $\ell$ is a decision
literal then $\reason{\ell} = \varnothing$.

In most \sat solver implementations clause learning is initiated as
soon as a clause is falsified by $\trail$. In this paper we will be
concerned with the subsequent clause learning process which uses
$\trail$ to derive a new clause that can be added to the formula. In
some implementations the trail might be altered during clause
learning. Here, however, we will assume that $\trail$ remains intact
during clause learning and is only changed after the new clause is
derived. After the new clause is lerned the $\trail$ will be changed by backtracking.

Say that $\trail$ falsifies a clause $c_I$, and that the last decision
literal $d_k$ in $\trail$ has decision level $k$. Consider
$\trail_{k-1}$ the prefix of $\trail$ above the last decision level,
i.e., the sequence of literals
$\trail[0]$---$\trail[\trailIdx{d_k}-1]$. We will assume that
$\trail_{k-1}$ is \textbf{propagation complete}, although the full
trail $\trail$ might not be. This means that (a) no clause was
falsified by $\trail_{k-1}$. And (b) if $c_u$ is a clause containing
the literal $x$ and all literals in $c_u$ except for $x$ are falsified
by $\trail_{k-1}$, then $x\in \trail_{k-1}$ and
$\dlevel{x} \leq\max\{\dlevel{y} | y\in c_u \land y\neq x\}$. This
means that if $x$ appears in a clause made unit it must have been
added to the trail, and added at the first decision level that this
occurred.  Note, however, $x$ might appear in more than one clause.
The less than equal condition above ensures that $x$ must appear in
$\trail_{k-1}$ at the first decision level any of these clauses were
made unit.

Any clause falsified by $\trail$ is called a \textbf{conflict}. When a
conflict is found, the final level of the trail, $k$, need not be
propagation complete as the solver typically stops propagation as soon
as it finds a conflict. This means that (a) other clauses might be
falsified by $\trail$ besides the conflict found, and (b) other
literals might be unit implied by $\trail$ but not added to $\trail$.

\begin{defn}[Trail Resolvant]
    A trail resolvant is a clause arising from resolving a conflict
    against the reason clause of some literal $\ell \in \trail$ where
    $\reason{\ell}\neq \varnothing$. Every trail resolvant is also a
    conflict.
\end{defn}

The following things can be noted about trail resolvants: (1) trail
resolvants are never tautological, as the polarity of all literals in
$\reason{\ell}$ other than $\ell$ must agree with the polarity of all
literals in the conflict (they are all falsified by $\trail$); (2) one
polarity of the variable $\var(\ell)$ resolved on must be a unit
propagated literal whose negation appears in the conflict; and (3) any
literal in the conflict that is unit progagated in $\trail$ can be
resolved upon (the corresponding variable must appear in different
polarities in the conflict and in $\trail$).

\begin{defn}[Trail Resolution]
    A trail resolution is a sequence of trail resolvants applied to an
    initial conflict $c_I$ yielding a new conflict $c_L$.  A trail
    resolution is \textbf{ordered} if the sequence of variables $v_1$,
    \dots, $v_m$ resolved have strictly decreasing trail indicies:
    $\trailIdx{v_i} > \trailIdx{v_{i+1}}$ ($1\leq i < m$). (Note that
    this implies that no variable is resolved on more than once).
\end{defn}

Ordered trail resolutions resolve unit propagated literals from the
end of the trail to the beginning. Without lost of generality, we can
require that all trail resolutions be ordered.

\begin{Obs}
    If the unordered trail resolution $U$ yields the conflict clause
    $c_L$ from an initial conflict $c_I$, then there exists an ordered
    trail resolution $O$ that yields a conflict clause $c'_L$ such
    that $c'_L\subseteq c_L$.
\end{Obs}
\noindent
\textbf{Proof.} Let $U$ be the sequence of clauses $c_I = c_0$, $c_1$,
$\ldots$, $c_m= c_L$ obtained by resolving on the sequence of
variables $v_1$, $\ldots$, $v_m$ whose corresponding literals on
$\trail$ are $l_1$, $\ldots$, $l_m$. Reordering these resolution steps
so that the variables are resolved in order of decreasing trail index
and removing duplicates yields an ordered trail resolution $O$ with
the desired properties. Since no reason clause contains literals with
higher trail indicies, $O$ must be a valid trail resolution if $U$
was, and furthermore $O$ yields the clause
$c'_L = \bigcup_{i=1}^m \reason{l_i} - \{l_1, \lnot l_1, \ldots, l_m,
\lnot l_m\}$. Since $U$ resolves on the same variables (in a different
order) we must that that the clause it produces $c_L\subseteq
c'_L$. But $c_L$ might retain one or more of the literals $\lnot l_1$,
$\ldots$, $\lnot l_m$, since resolving on a literal $l_j$ with higher
trail index might reintroduce a literal $\lnot l_i$ with lower trail
index that was previously resolved away by an earlier resolution.
\whitebox

The relevance of trail resolutions is that all proposed clause
learning schemes we are aware of use trail resolutions to produce
learnt clauses. Furthermore, the commonly used technique for clause
minimization \cite{DBLP:conf/sat/SorenssonB09} is also equivalent to a
trail resolution that yields the minimized clause from the
un-minimized clause. Interestingly, it is standard in \sat solver
implementations to perform resolution going backwards along the
trail. That is, these implementations are typically using ordered
trail resolutions. The above observation shows that this in fact the
right way to do this.

Ordered trail resolutions are a special case of \textit{trivial
  resolutions} \cite{DBLP:journals/jair/BeameKS04}. Trail resolutions
are specific to the trail data structure typically used in \sat
solvers. If $\trail$ falsifies a clause at its last decision level,
then its associated implication graph
\cite{DBLP:journals/tc/Marques-SilvaS99} contains a conflict
node. Cuts in the implication graph that separate the conflict from
the rest of the graph correspond to conflict clauses
\cite{DBLP:journals/jair/BeameKS04}. It is not difficult to see that
the proof Proposition~4 of \cite{DBLP:journals/jair/BeameKS04} applies
also to trail resolutions. This means that \textit{any conflict clause
  in the trail's implication graph can be derived using a trail
  resolution.}

\subsection{Some alternate Clause Learning Schemes}
A number of different clause learning schemes for generating a new
learnt clause from the initial conflict have been presented in prior
work, e.g.,
\cite{DBLP:conf/iccad/ZhangMMM01,DBLP:conf/iccad/SilvaS96,DBLP:journals/tc/Marques-SilvaS99,DBLP:conf/aaai/BayardoS97}.
Figure~\ref{fig:cl_schemes} gives a specification of some of these
methods: (a) the all-decision scheme which resolves away all implied
literals leaving a learnt clause over only decision literals; (c) the
1-UIP scheme which resolves away literals from the deepest decision
level leaving a learnt clause with a single literal at the deepest
level; (d) the all-UIP scheme which resolves away literals from each
decision level leaving a learnt clause with a single literal at each
decision level; and (e) the $i$-UIP scheme which resolves away
literals from the $i$ deepest decision levels leaving a learnt clause
with a single literal at its $i$ deepest decision levels. It should be
noted that when resolving away literals at decision level $i$ new
literals at decision levels less than $i$ might be introduced into the
clause. Hence, it is important in the $i$-UIP and all-UIP schemes to
use ordered trail resolutions.

Both the all-decision and all-UIP schemes yield a clause with only one
literal at each decision level, and the all-UIP clause will be no
larger that the all-decision clause. Furthermore, it is known
\cite{DBLP:journals/tc/Marques-SilvaS99} that once we reduce the
number of literals at a decision level $d$ to one, we could continue
performing resolutions and later achieve a different single literal at
the level $d$. In particular, a decision level might contain more than
one unique implication point. The algorithms given in
Figure~\ref{fig:cl_schemes} stop at the first UIP of a level, except
for the all-decision schemes with stops at the last UIP of each level.
